---
title: "CPP-528 Lab 04"
author: "Sarah Johaningsmeir"
date: '2022-04-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}

library( dplyr )
library( here )
library( knitr )
library( pander )
library( stargazer )
library( scales )

# set randomization seed ----
set.seed( 1234 )

import::here("S_TYPE",
             "panel.cor",
             "panel.smooth",
             "jplot",
             "d",
             "df",
             "cbsa_stats_df",
             .from = here::here("labs/wk04/lab_04_source.R"),
             .character_only = TRUE)

```


## Part 1, Data

```{r}

import::here("clean_d",
             "tidy_up_data",
             "build_year",
             "RELEVANT_FILES",
             "obtain_crosswalk",
             "create_final_metadata_file",
             # notice the use of here::here() that points to the .R file
             # where all these R objects are created
             .from = here::here("labs/wk03/utilities-2.R"),
             .character_only = TRUE)

```



```{r}
d1 <- readRDS( here::here( "data/rodeo/LTDB-2000.rds" ) )
d2 <- readRDS( here::here( "data/rodeo/LTDB-2010.rds" ) )
md <- readRDS( here::here( "data/rodeo/LTDB-META-DATA.rds" ) )

```

```{r}

d1 <- select( d1, - year )
d2 <- select( d2, - year )

d <- merge( d1, d2, by="tractid" )
d <- merge( d, md, by="tractid" )

```


### Filter to just urban tracts 

```{r}

table( d$urban )

d <- filter( d, urban == "urban" )

```

```{r}


# find variables that are in both files
compare_dfs <- function( df1, df2 )
{
  # use regular expressions to remove numeric suffixes 
  var.names.1 <- names( df1 )
  var.names.1 <- gsub( "[.][xy]$", "", var.names.1 )
  var.names.1 <- gsub( "[0-9]{2}$", "", var.names.1 )
  
  var.names.2 <- names( df2 )
  var.names.2 <- gsub( "[.][xy]$", "", var.names.2 )
  var.names.2 <- gsub( "[0-9]{2}$", "", var.names.2 )
  
  shared <- intersect( var.names.1, var.names.2 ) %>% sort()
  print( "SHARED VARIABLES:")
  print( shared )
  
  not.shared <- c( setdiff( var.names.1, var.names.2 ),
                   setdiff( var.names.2, var.names.1 ) ) %>% sort()
  
  print( "NOT SHARED:" )
  print( not.shared )
  
  d.vars1 <- data.frame( type="shared", variables=shared, stringsAsFactors=F )
  d.vars2 <- data.frame( type="not shared", variables=not.shared, stringsAsFactors=F )
  dd <- rbind( d.vars1, d.vars2 )
  
  return( dd )
}

vars <- compare_dfs( df1=d1, df2=d2 )

```

```{r}

d.full <- d  

```



### 2000-2010 variables

```{r}

d <- d.full  

d <- select( d, tractid, mhmval00, mhmval12, hinc00, 
             hu00, own00, rent00,  
             empclf00, clf00, unemp00, prof00,  
             dpov00, npov00,
             ag25up00, hs00, col00, 
             pop00.x, nhwht00, nhblk00, hisp00, asian00,
             cbsa, cbsaname )
d <- 
  d %>%
  mutate( p.white = 100 * nhwht00 / pop00.x,
          p.black = 100 * nhblk00 / pop00.x,
          p.hisp = 100 * hisp00 / pop00.x, 
          p.asian = 100 * asian00 / pop00.x,
          p.hs = 100 * (hs00+col00) / ag25up00,
          p.col = 100 * col00 / ag25up00,
          p.prof = 100 * prof00 / empclf00,
          p.unemp = 100 * unemp00 / clf00,
          pov.rate = 100 * npov00 / dpov00 )

```



```{r, results='asis'}

# adjust 2000 home values for inflation 
mhv.00 <- d$mhmval00 * 1.28855  
mhv.10 <- d$mhmval12

mhv.change <- mhv.10 - mhv.00

df <- data.frame( MedianHomeValue2000=mhv.00, 
                  MedianHomeValue2010=mhv.10, 
                  Change.00.to.10=mhv.change )

stargazer( df, 
           type="html", 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )

```


### Change in MHV 2000-2010

```{r}

#filter out properties valued under $10,000 in 2000

mhv.00[ mhv.00 < 10000 ] <- NA
pct.change <- mhv.change / mhv.00
summary( pct.change )

```

### Group growth rates by metro area

```{r}

d$mhv.change <- mhv.change 
d$pct.change <- pct.change
d$mhv.10 <- mhv.10
d$mhv.00 <- mhv.00


```


### Select gentrification variables

```{r}

d.full$mhv.00 <- mhv.00
d.full$mhv.10 <- mhv.10
d.full$mhv.change <- mhv.change
d.full$pct.change <- pct.change

```




```{r}

d3 <- select( d.full, 
             
             tractid, cbsa, cbsaname,            # ids / units of analysis
             
             mhv.00, mhv.10, mhv.change, pct.change,    # home value 
             
             hinc00, hu00, own00, rent00,        # ses
             hinc12, hu10, own10, rent10,
             
             empclf00, clf00, unemp00, prof00,   # employment 
             empclf12, clf12, unemp12, prof12,
             
             dpov00, npov00,                     # poverty
             dpov12, npov12,
             
             ag25up00, hs00, col00,              # education 
             ag25up12, hs12, col12,
             
             pop00.x, nhwht00, nhblk00, hisp00, asian00,   # race
             pop10, nhwht10, nhblk10, hisp10, asian10
             
          ) # end select

d3 <- 
  d3 %>%
  mutate( 
          # 2000 variables
          p.white.00 = 100 * nhwht00 / pop00.x,
          p.black.00 = 100 * nhblk00 / pop00.x,
          p.hisp.00 = 100 * hisp00 / pop00.x, 
          p.asian.00 = 100 * asian00 / pop00.x,
          p.hs.edu.00 = 100 * (hs00+col00) / ag25up00,
          p.col.edu.00 = 100 * col00 / ag25up00,
          p.prof.00 = 100 * prof00 / empclf00,
          p.unemp.00 = 100 * unemp00 / clf00,
          pov.rate.00 = 100 * npov00 / dpov00,
          
          # 2010 variables
          p.white.10 = 100 * nhwht10 / pop10,
          p.black.10 = 100 * nhblk10 / pop10,
          p.hisp.10 = 100 * hisp10 / pop10, 
          p.asian.10 = 100 * asian10 / pop10,
          p.hs.edu.10 = 100 * (hs12+col12) / ag25up12,
          p.col.edu.10 = 100 * col12 / ag25up12,
          p.prof.10 = 100 * prof12 / empclf12,
          p.unemp.10 = 100 * unemp12 / clf12,
          pov.rate.10 = 100 * npov12 / dpov12 )

```


```{r}

d3 <-
  d3 %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.pct.00 = ntile( mhv.00, 100 ),
          metro.mhv.pct.10 = ntile( mhv.10, 100 ),
          metro.median.pay.00 = median( hinc00, na.rm=T ),
          metro.median.pay.10 = median( hinc12, na.rm=T ),
          metro.race.rank.00 = ntile( (100-p.white.00), 100 ) ) %>%
  ungroup() %>%
  mutate( metro.mhv.pct.change = metro.mhv.pct.10 - metro.mhv.pct.00,
          pay.change = metro.median.pay.10 - metro.median.pay.00,
          race.change = p.white.10 - p.white.00,
          mhv.change = mhv.10 - mhv.00 )

```


### Descriptive statistics of change variables

```{r, results='asis'}

d3 <-           
  d3 %>%
  select( c( "tractid", "cbsa", "cbsaname",
             "mhv.00", "mhv.10", "mhv.change","pct.change",
          "p.white.00", "p.black.00", "p.hisp.00", "p.asian.00", 
          "p.hs.edu.00", "p.col.edu.00", "p.prof.00",  "p.unemp.00", 
          "pov.rate.00", "p.white.10", "p.black.10", "p.hisp.10", 
          "p.asian.10", "p.hs.edu.10", "p.col.edu.10", "p.prof.10", 
          "p.unemp.10", "pov.rate.10", "metro.mhv.pct.00", 
          "metro.mhv.pct.10", "metro.median.pay.00", "metro.median.pay.10", 
          "metro.mhv.pct.change", "pay.change", "race.change",
          "metro.race.rank.00") ) 
  
d3 <- data.frame(d3)
stargazer( d3, 
           type="html", 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )

```


### Remove tracts where properties were worth less than $10,000 in 2000

```{r}

mhv.00[ mhv.00 < 10000 ] <- NA
pct.change <- 100 * ( mhv.change / mhv.00 )
summary( pct.change )

```

### Remove tracts with growth above 200%

```{r}

pct.change[ pct.change > 200 ] <- NA

```


# Part 2, Predict MHV Change


## Select at least three census variables that you feel will be good predictors of change in MHV between 2000 and 2010. 

* Percent white residents in 2000 (p.white.00)
* Percent with 4-year college education or more (p.col.edu.00)
* Poverty rate in 2000 (pov.rate.00)


## Check for variable skew 

```{r}

hist( d3$p.white.00, breaks = 50 )

hist( log( d3$p.white.00 + 1), breaks = 50 )

hist( d3$p.col.edu.00, breaks = 50 )

hist( log( d3$p.col.edu.00 + 1 ), breaks = 50 )

hist( d3$pov.rate.00, breaks = 50 )

hist( log( d3$pov.rate.00 + 1 ), breaks = 50 )

```


```{r}

log.p.white.00 <- log10( d3$p.white.00 + 1 )
log.p.col.edu.00 <- log10( d3$p.col.edu.00 + 1 )

par( mfrow=c( 1, 2 ))

these <- sample( 1:length( log.p.white.00), 5000 )

jplot( d3$p.white.00[these], d3$p.col.edu.00[these],
       lab1 = "Percent White", lab2 = "Percent College Educated",
       main = "Raw Measures" )

jplot( log.p.white.00[these], log.p.col.edu.00[these],
       lab1 = "Percent White", lab2 = "Percent College Educated",
       main = "Log Transformed" )

```

```{r}

cor( d3$p.white.00, d3$p.col.edu.00, use = "pairwise.complete" )

cor( log.p.white.00, log.p.col.edu.00, use = "pairwise.complete" )

```


```{r}

hist( pct.change, breaks = 50 )

hist( mhv.00, breaks = 50 )

hist( log( mhv.00 + 1 ), breaks = 50 )

hist( mhv.change, breaks = 50 )

hist( log( mhv.change + 1 ), breaks = 50 )

```

Determined that logging improves the skew for p.white.00, p.col.edu.00, pov.rate.00,
mhv.00, and mhv.change, but not pct.change. 


## Check for multicollinearity  


```{r}
# Correlations of raw scores

d4 <- select( d3, p.white.00, p.col.edu.00, pov.rate.00, mhv.00, mhv.change, pct.change )

set.seed( 1234 )
d4 <- sample_n( d4, 10000 ) %>% na.omit()

pairs( d4, upper.panel=panel.cor, lower.panel=panel.smooth )

```

```{r}
# Correlations of logged scores

set.seed( 1234 )

d4 <- select( d3, p.white.00, p.col.edu.00, pov.rate.00, mhv.change )

d4$p.white.00 <- log10( d4$p.white.00 + 1 )
d4$p.col.edu.00 <- log10( d4$p.col.edu.00 + 1 )
d4$pov.rate.00 <- log10( d4$pov.rate.00 + 1 )
d4$mhv.change <- log10( d4$mhv.change + 1 )

d5 <- sample_n( d4, 5000 ) %>% na.omit()

pairs( d5, upper.panel = panel.cor, lower.panel = panel.smooth )


```


## Regression models with metro-area fixed effects


```{r}

d.reg <- d

d.reg$mhv.change[ d.reg$mhv.change > 200 ] <- NA
d.reg$p.white <- log10( d.reg$p.white + 1 )
d.reg$p.col <- log10( d.reg$p.col + 1 )
d.reg$pov.rate <- log10( d.reg$pov.rate + 1 )

# average growth in median home value for the city
d.reg <- 
  d.reg %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.growth = 100 * median( mhv.change, na.rm=T ) ) %>%
  ungroup() 

```


### Each variable alone (with metro-area fixed effects)

```{r, results='asis'}

m1 <- lm( mhv.change ~ p.white + cbsa, data = d.reg )
m2 <- lm( mhv.change ~ p.col + cbsa, data = d.reg )
m3 <- lm( mhv.change ~ pov.rate + cbsa, data = d.reg )

stargazer( m1, m2, m3,
           type = "html", digits = 2,
           omit.stat = c( "rsq", "f" ),
           omit = "cbsa" )

```

### Regression with multiple variables

```{r, results='asis'}


m1 <- lm( mhv.change ~ p.white + p.col + cbsa, data=d.reg )
m2 <- lm( mhv.change ~ p.white + pov.rate + cbsa, data = d.reg)
m3 <- lm( mhv.change ~ pov.rate + p.col + cbsa, data = d.reg )
m4 <- lm( mhv.change ~ p.white + p.col + pov.rate + cbsa, data=d.reg)

stargazer( m1, m2, m3, m4,  
           type= "html", digits=2,
           omit.stat = c("rsq","f"),
           omit="cbsa" )

```


## Part 4 Results

Each of the three variables I chose (percent white residents, percent with college
or greater education level, and poverty rate) did significantly predict change in
median home values from 2000 to 2010. Controlling for metro area fixed effects, the 
amount of variance accounted for was approximately equal whether looking at any 
of these variables alone or in a model including two or three variable--each of
these models had an Adjusted R2 of 0.29 to 0.31. This is likely because of the strong
correlations between the variables. 

Model 4 with all three variables included produced an Adjusted R2 of 0.31, 
significant  at the p<0.01 level. Each of these variables had a negative correlation,
thus a higher percentage of white residents, higher percentage with a college
education, or higher poverty rate all indicate a smaller change in median home values.

I expected to see these relationships for percent of white and college-educated 
residents, since tracts with these demographics may already be high-priced and not
eligible for gentrification (and I also expect the converse, that areas with low 
white and college-educated populations may be ripe for gentrification). However, I
am surprised to see that poverty rate is also a negative correlation. This may be due 
to multicollinearity, since college education and poverty rate are closely correlated. 
In Model 2 where college education is not included, poverty rate has a positive
correlation with median home value change.







