---
title: "Lab 04 - Predicting MHV Change"
author: "Mohamed Said"
date: '2022-04-9'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```


```{r}
# load necessary packages ----
library( dplyr )
library( here )
library( knitr )
library( pander )
library( stargazer )
library( scales )

# set randomization seed ----
set.seed( 1234 )

# load necessary functions and objects ----
# note: all of these are R objects that will be used throughout this .rmd file
import::here("S_TYPE",
             "panel.cor",
             "panel.smooth",
             "jplot",
             "d",
             "df",
             "cbsa_stats_df",
             # notice the use of here::here() that points to the .R file
             # where all these R objects are created
             .from = here::here("labs/wk04/lab_04_source.R"),
             .character_only = TRUE)
```

*Part-One-Data*

```{r}
# note: please do not use static file paths
# note: notice down below the use of here::here()
d1 <- readRDS( here::here( "data/rodeo/LTDB-2000.rds" ) )
d2 <- readRDS( here::here( "data/rodeo/LTDB-2010.rds" ) )
md <- readRDS( here::here( "data/rodeo/LTDB-META-DATA.rds" ) )
# check to make sure we are not losing 
# or gaining observations in the merge
nrow( d1 ) 
```

```{r}
d1 <- select( d1, - year )
d2 <- select( d2, - year )
d <- merge( d1, d2, by="tractid" )
d <- merge( d, md, by="tractid" )
nrow( d )
```

**Filter Rural Districts**

```{r}
#urban districts only
table( d$urban )
```

```{r}
d <- filter( d, urban == "urban" )
```

*Identify Common Variables*

```{r}
# find common variables in both files
compare_dfs <- function( df1, df2 )
{
  # use regular expressions to remove numeric suffixes 
  var.names.1 <- names( df1 )
  var.names.1 <- gsub( "[.][xy]$", "", var.names.1 )
  var.names.1 <- gsub( "[0-9]{2}$", "", var.names.1 )
  
  var.names.2 <- names( df2 )
  var.names.2 <- gsub( "[.][xy]$", "", var.names.2 )
  var.names.2 <- gsub( "[0-9]{2}$", "", var.names.2 )
  
  shared <- intersect( var.names.1, var.names.2 ) %>% sort()
  print( "SHARED VARIABLES:")
  print( shared )
  
  not.shared <- c( setdiff( var.names.1, var.names.2 ),
                   setdiff( var.names.2, var.names.1 ) ) %>% sort()
  
  print( "NOT SHARED:" )
  print( not.shared )
  
  d.vars1 <- data.frame( type="shared", variables=shared, stringsAsFactors=F )
  d.vars2 <- data.frame( type="not shared", variables=not.shared, stringsAsFactors=F )
  dd <- rbind( d.vars1, d.vars2 )
  
  return( dd )
}
vars <- compare_dfs( df1=d1, df2=d2 )
```



```{r}
head( vars )
```

**Create data set for analysis**

```{r}
d.full <- d  # keep a copy so don't have to reload 
```

**2000-2010 variables**

```{r}
d <- d.full  # store original in case you need to reset anything
d <- select( d, tractid, 
             mhmval00, mhmval12, 
             hinc00, 
             hu00, vac00, own00, rent00, h30old00,
             empclf00, clf00, unemp00, prof00,  
             dpov00, npov00,
             ag25up00, hs00, col00, 
             pop00.x, nhwht00, nhblk00, hisp00, asian00,
             cbsa, cbsaname)

d <- 
  d %>%
  dplyr::mutate( p.white = 100 * nhwht00 / pop00.x,
          p.black = 100 * nhblk00 / pop00.x,
          p.hisp = 100 * hisp00 / pop00.x, 
          p.asian = 100 * asian00 / pop00.x,
          p.hs = 100 * (hs00+col00) / ag25up00,
          p.col = 100 * col00 / ag25up00,
          p.prof = 100 * prof00 / empclf00,
          p.unemp = 100 * unemp00 / clf00,
          p.vacant = 100 * vac00 / hu00,
          mhv.change.00.to.10 = mhmval12 - mhmval00,
          p.mhv.change = 100 * (mhmval12 - mhmval00) / mhmval00,
          pov.rate = 100 * npov00 / dpov00 )

head(d)
```


**Exploration of median home value**

```{r}
# adjust 2000 home values for inflation 
mhv.00 <- d$mhmval00 * 1.28855  
mhv.10 <- d$mhmval12
mhv.change <- mhv.10 - mhv.00
df <- data.frame( MedianHomeValue2000=mhv.00, 
                  MedianHomeValue2010=mhv.10, 
                  Change.00.to.10=mhv.change )
stargazer( df, 
           type="text", 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```


```{r}
# Plot the percent change variable
hg <-
hist( pct.change, breaks=2000, 
      xlim=c(-1,2), yaxt="n", xaxt="n",
      xlab="", cex.main=1.5,
      ylab="", main="Growth in Home Value by Census Tract 2000 to 2010",
      col="gray40", border="white" )
axis( side=1, at=seq( from=-1, to=2, by=0.5 ), 
      labels=paste0( seq( from=-100, to=200, by=50 ), "%" ) )
ymax <- max( hg$count )
        
mean.x <- mean( pct.change, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=1, y=(0.5*ymax), 
      labels=paste0( "Mean = ", round(100*mean.x,0), "%"), 
      col="darkorange", cex=1.8, pos=4 )
median.x <- median( pct.change, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=1, y=(0.6*ymax), 
      labels=paste0( "Median = ", round(100*median.x,0), "%"), 
      col="dodgerblue", cex=1.8, pos=4 )
```

```{r}
hist( mhv.change/1000, breaks=500, 
      xlim=c(-100,500), yaxt="n", xaxt="n",
      xlab="Thousand of US Dollars (adjusted to 2010)", cex.lab=1.5,
      ylab="", main="Change in Median Home Value 2000 to 2010",
      col="gray20", border="white" )
axis( side=1, at=seq( from=-100, to=500, by=100 ), 
      labels=paste0( "$", seq( from=-100, to=500, by=100 ), "k" ) )
        
mean.x <- mean( mhv.change/1000, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=200, y=1500, 
      labels=paste0( "Mean = ", dollar( round(1000*mean.x,0)) ), 
      col="darkorange", cex=1.8, pos=3 )
median.x <- median( mhv.change/1000, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=200, y=2000, 
      labels=paste0( "Median = ", dollar( round(1000*median.x,0)) ), 
      col="dodgerblue", cex=1.8, pos=3 )
```
*Omit cases that have a median home value less than $10,000 in 2000.*
```{r}
# quick filter to remove all of the problematic obs
# but need to go back and see which cases are problematic
mhv.00[ mhv.00 < 10000 ] <- NA

pct.change <- mhv.change / mhv.00
summary( pct.change )
```

```{r}
# how many cases had increases above 200%
sum( pct.change > 2, na.rm=T )
```

```{r}
# preview tracts with large increases in home values 
# to see if increases make sense 
d %>% 
  filter( pct.change > 2 ) %>% 
  head()
```

*Growth Rate*

```{r}
d$mhv.change <- mhv.change 
d$pct.change <- pct.change
d$mhv.10 <- mhv.10
d$mhv.00 <- mhv.00
```


### Select gentrification variables

```{r}
d.full$mhv.00 <- mhv.00
d.full$mhv.10 <- mhv.10
d.full$mhv.change <- mhv.change
d.full$pct.change <- pct.change
```




```{r}
d3 <- select( d.full, 
             
             tractid, cbsa, cbsaname,            # ids / units of analysis
             
             mhv.00, mhv.10, mhv.change, pct.change,    # home value 
             
             hinc00, hu00, own00, rent00,        # ses
             hinc12, hu10, own10, rent10,
             
             empclf00, clf00, unemp00, prof00,   # employment 
             empclf12, clf12, unemp12, prof12,
             
             dpov00, npov00,                     # poverty
             dpov12, npov12,
             
             ag25up00, hs00, col00,              # education 
             ag25up12, hs12, col12,
             
             pop00.x, nhwht00, nhblk00, hisp00, asian00,   # race
             pop10, nhwht10, nhblk10, hisp10, asian10
             
          ) # end select
d3 <- 
  d3 %>%
  mutate( 
          # 2000 variables
          p.white.00 = 100 * nhwht00 / pop00.x,
          p.black.00 = 100 * nhblk00 / pop00.x,
          p.hisp.00 = 100 * hisp00 / pop00.x, 
          p.asian.00 = 100 * asian00 / pop00.x,
          p.hs.edu.00 = 100 * (hs00+col00) / ag25up00,
          p.col.edu.00 = 100 * col00 / ag25up00,
          p.prof.00 = 100 * prof00 / empclf00,
          p.unemp.00 = 100 * unemp00 / clf00,
          pov.rate.00 = 100 * npov00 / dpov00,
          
          # 2010 variables
          p.white.10 = 100 * nhwht10 / pop10,
          p.black.10 = 100 * nhblk10 / pop10,
          p.hisp.10 = 100 * hisp10 / pop10, 
          p.asian.10 = 100 * asian10 / pop10,
          p.hs.edu.10 = 100 * (hs12+col12) / ag25up12,
          p.col.edu.10 = 100 * col12 / ag25up12,
          p.prof.10 = 100 * prof12 / empclf12,
          p.unemp.10 = 100 * unemp12 / clf12,
          pov.rate.10 = 100 * npov12 / dpov12 )
```


```{r}
d3 <-
  d3 %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.pct.00 = ntile( mhv.00, 100 ),
          metro.mhv.pct.10 = ntile( mhv.10, 100 ),
          metro.median.pay.00 = median( hinc00, na.rm=T ),
          metro.median.pay.10 = median( hinc12, na.rm=T ),
          metro.race.rank.00 = ntile( (100-p.white.00), 100 ) ) %>%
  ungroup() %>%
  mutate( metro.mhv.pct.change = metro.mhv.pct.10 - metro.mhv.pct.00,
          pay.change = metro.median.pay.10 - metro.median.pay.00,
          race.change = p.white.10 - p.white.00,
          mhv.change = mhv.10 - mhv.00 )
```


### Descriptive statistics of change variables

```{r}
d3 <-           
  d3 %>%
  select( c( "tractid", "cbsa", "cbsaname",
             "mhv.00", "mhv.10", "mhv.change","pct.change",
          "p.white.00", "p.black.00", "p.hisp.00", "p.asian.00", 
          "p.hs.edu.00", "p.col.edu.00", "p.prof.00",  "p.unemp.00", 
          "pov.rate.00", "p.white.10", "p.black.10", "p.hisp.10", 
          "p.asian.10", "p.hs.edu.10", "p.col.edu.10", "p.prof.10", 
          "p.unemp.10", "pov.rate.10", "metro.mhv.pct.00", 
          "metro.mhv.pct.10", "metro.median.pay.00", "metro.median.pay.10", 
          "metro.mhv.pct.change", "pay.change", "race.change",
          "metro.race.rank.00") ) 
d3 <- data.frame(d3)
stargazer( d3, 
           type="text", 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

*Remove tracts where properties were worth less than $1,000 in 2000*

```{r}
mhv.00[ mhv.00 < 1000 ] <- NA
pct.change <- 100 * ( mhv.change / mhv.00 )
summary( pct.change )
```

*Remove tracts with growth above 200%*

```{r}
pct.change[ pct.change > 200 ] <- NA

names(d3)

```


*Part 2 - Predict MHV Change*

**Select at least three census variables that you feel will be good predictors of change in MHV between 2000 and 2010.**


**Percent black residents in 2000 (p.white.00)**
**Percent with 4-year college education or more (p.col.edu.00)**
**Poverty rate in 2000 (pov.rate.00)**


*Check for variable skew*

```{r}
hist( d3$p.white.00, breaks = 50 )
hist( log( d3$p.white.00 + 1), breaks = 50 )
hist( d3$p.col.edu.00, breaks = 50 )
hist( log( d3$p.col.edu.00 + 1 ), breaks = 50 )
hist( d3$pov.rate.00, breaks = 50 )
hist( log( d3$pov.rate.00 + 1 ), breaks = 50 )
```

```{r}
p.white.00 <- log10( d3$p.white.00 + 1 )
log.p.col.edu.00 <- log10( d3$p.col.edu.00 + 1 )
par( mfrow=c( 1, 2 ))
these <- sample( 1:length( p.white.00), 5000 )
jplot( d3$p.white.00[these], d3$p.col.edu.00[these],
       lab1 = "Percent White", lab2 = "Percent College Educated",
       main = "Raw Measures" )
jplot( p.white.00[these], log.p.col.edu.00[these],
       lab1 = "Percent White", lab2 = "Percent College Educated",
       main = "Log Transformed" )
```
```{r}
cor( d3$p.white.00, d3$p.col.edu.00, use = "pairwise.complete" )
cor( p.white.00, log.p.col.edu.00, use = "pairwise.complete" )
```
```{r}
hist( pct.change, breaks = 50 )
hist( mhv.00, breaks = 50 )
hist( log( mhv.00 + 1 ), breaks = 50 )
hist( mhv.change, breaks = 50 )
hist( log( mhv.change + 1 ), breaks = 50 )
```

**the log improved the percent of education education, percent of white, poverty rate and , percent of change  but not Medium home value change**

**Check for multidisciplinary**


```{r}
# Correlations of raw scores
d4 <- select( d3, p.white.00, p.col.edu.00, pov.rate.00, mhv.00, mhv.change, pct.change )
set.seed( 1234 )
d4 <- sample_n( d4, 10000 ) %>% na.omit()
pairs( d4, upper.panel=panel.cor, lower.panel=panel.smooth )
```
*Correlations of logged scores*

```{r}
set.seed( 1234 )
d4 <- select( d3, p.white.00, p.col.edu.00, pov.rate.00, mhv.change )
d4$p.white.00 <- log10( d4$p.white.00 + 1 )
d4$p.col.edu.00 <- log10( d4$p.col.edu.00 + 1 )
d4$pov.rate.00 <- log10( d4$pov.rate.00 + 1 )
d4$mhv.change <- log10( d4$mhv.change + 1 )
d5 <- sample_n( d4, 5000 ) %>% na.omit()
pairs( d5, upper.panel = panel.cor, lower.panel = panel.smooth )
```


*Regression models with metro-area fixed effects*


```{r}
d.reg <- d
d.reg$mhv.change[ d.reg$mhv.change > 200 ] <- NA
d.reg$p.white <- log10( d.reg$p.white + 1 )
d.reg$p.col <- log10( d.reg$p.col + 1 )
d.reg$pov.rate <- log10( d.reg$pov.rate + 1 )
# average growth in median home value for the city
d.reg <- 
  d.reg %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.growth = 100 * median( mhv.change, na.rm=T ) ) %>%
  ungroup() 
```


*Each variable alone (with metro-area fixed effects)*

```{r}
m1 <- lm( mhv.change ~ p.white + cbsa, data = d.reg )
m2 <- lm( mhv.change ~ p.col + cbsa, data = d.reg )
m3 <- lm( mhv.change ~ pov.rate + cbsa, data = d.reg )
stargazer( m1, m2, m3,
           type = "text", digits = 2,
           omit.stat = c( "rsq", "f" ),
           omit = "cbsa" )
```

*Regression with multiple variables*

```{r}
m1 <- lm( mhv.change ~ p.white + p.col + cbsa, data=d.reg )
m2 <- lm( mhv.change ~ p.white + pov.rate + cbsa, data = d.reg)
m3 <- lm( mhv.change ~ pov.rate + p.col + cbsa, data = d.reg )
m4 <- lm( mhv.change ~ p.white + p.col + pov.rate + cbsa, data=d.reg)
stargazer( m1, m2, m3, m4,  
           type= "text", digits=2,
           omit.stat = c("rsq","f"),
           omit="cbsa" )
```


**All the selected Variables successfully predicted the medium home value change form 2000 to 2010. There was significant relationship between the selected for you please that’s why the R2 was around to 0.3 But adding another variables but like unemployment will have a strong relationship with the median home value**

**I feel that there will be difference in the results if we use different demographics like percent of black and Asians.**







