---
title: "CPP-528 Lab 04"
author: "Team 2"
date: '2022-04-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, echo=FALSE, include=FALSE, warning=FALSE}

library( dplyr )
library( here )
library( kableExtra )
library( knitr )
library( pander )
library( stargazer )
library( scales )

# set randomization seed ----
set.seed( 1234 )

import::here("S_TYPE",
             "panel.cor",
             "panel.smooth",
             "jplot",
             "d",
             "df",
             "d.full",
             "d3",
             "pct.change",
             .from = here::here("labs/wk06/lab_06_source.R"),
             .character_only = TRUE)

```


## Part 1, Data


### Summary statistics of Median Home Value in 2000, 2010, and change

```{r, results='asis'}

stargazer( df, 
           type="html", 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )

```


### Change in MHV 2000-2010

```{r, echo=FALSE}

summary( pct.change )

```


### Descriptive statistics of change variables

```{r, echo=FALSE, results='asis'}

df <- data.frame(df)
stargazer( df, 
           type="html", 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )

```

# Part 2, Predict MHV Change


## Select at least three census variables that you feel will be good predictors of change in MHV between 2000 and 2010. 

* Percent white residents in 2000 (p.white.00)
* Percent with 4-year college education or more (p.col.edu.00)
* Poverty rate in 2000 (pov.rate.00)


## Descriptive statistics of change variables

```{r, results='asis', echo=FALSE}

d3 <- data.frame(d3)
stargazer( d3, 
           type="html", 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )

```



## Check for variable skew
If variables are skewed, log to establish normal distribution

```{r, echo=FALSE}
par( mfrow=c(1,2) )

hist( d3$p.white.00, breaks = 50 )

hist( log( d3$p.white.00 + 1), breaks = 50 )

hist( d3$p.col.edu.00, breaks = 50 )

hist( log( d3$p.col.edu.00 + 1 ), breaks = 50 )

hist( d3$pov.rate.00, breaks = 50 )

hist( log( d3$pov.rate.00 + 1 ), breaks = 50 )

```


```{r, echo=FALSE}

log.p.white.00 <- log10( d3$p.white.00 + 1 )
log.p.col.edu.00 <- log10( d3$p.col.edu.00 + 1 )

par( mfrow=c( 1, 2 ))

these <- sample( 1:length( log.p.white.00), 5000 )

jplot( d3$p.white.00[these], d3$p.col.edu.00[these],
       lab1 = "Percent White", lab2 = "Percent College Educated",
       main = "Raw Measures" )

jplot( log.p.white.00[these], log.p.col.edu.00[these],
       lab1 = "Percent White", lab2 = "Percent College Educated",
       main = "Log Transformed" )

```

```{r, echo=FALSE, results='hide'}

cor( d3$p.white.00, d3$p.col.edu.00, use = "pairwise.complete" )

cor( log.p.white.00, log.p.col.edu.00, use = "pairwise.complete" )

```


```{r, warning=FALSE, echo=FALSE}

hist( pct.change, breaks = 50 )

par( mfrow=c(1,2) )

hist( d$mhv.00, breaks = 50 )

hist( log( d$mhv.00 + 1 ), breaks = 50 )

hist( d$mhv.change, breaks = 50 )

hist( log( d$mhv.change + 1 ), breaks = 50 )

```

Determined that logging improves the skew for p.white.00, p.col.edu.00, pov.rate.00,
mhv.00, and mhv.change, but not pct.change. 


## Check for multicollinearity  

### Correlations of raw scores

```{r, echo=FALSE}
# Correlations of raw scores

d4 <- select( d3, p.white.00, p.col.edu.00, pov.rate.00, mhv.00, mhv.change, pct.change )

set.seed( 1234 )
d4 <- sample_n( d4, 10000 ) %>% na.omit()

pairs( d4, upper.panel=panel.cor, lower.panel=panel.smooth )

```

### Correlations of logged scores

```{r, warning=FALSE, echo=FALSE}
# Correlations of logged scores

set.seed( 1234 )

d4 <- select( d3, p.white.00, p.col.edu.00, pov.rate.00, mhv.change )

d4$p.white.00 <- log10( d4$p.white.00 + 1 )
d4$p.col.edu.00 <- log10( d4$p.col.edu.00 + 1 )
d4$pov.rate.00 <- log10( d4$pov.rate.00 + 1 )
d4$mhv.change <- log10( d4$mhv.change + 1 )

d5 <- sample_n( d4, 5000 ) %>% na.omit()

pairs( d5, upper.panel = panel.cor, lower.panel = panel.smooth )


```


## Regression models with metro-area fixed effects


```{r, echo=FALSE}

d.reg <- d

d.reg$mhv.change[ d.reg$mhv.change > 200 ] <- NA
d.reg$p.white <- log10( d.reg$p.white.00 + 1 )
d.reg$p.col <- log10( d.reg$p.col.edu.00 + 1 )
d.reg$pov.rate <- log10( d.reg$pov.rate.00 + 1 )

# average growth in median home value for the city
d.reg <- 
  d.reg %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.growth = 100 * median( mhv.change, na.rm=T ) ) %>%
  ungroup() 

```


### Each variable alone (with metro-area fixed effects)

```{r, results='asis', echo=FALSE}

m1 <- lm( mhv.change ~ p.white + cbsa, data = d.reg )
m2 <- lm( mhv.change ~ p.col + cbsa, data = d.reg )
m3 <- lm( mhv.change ~ pov.rate + cbsa, data = d.reg )

stargazer( m1, m2, m3,
           type = "html", digits = 2,
           omit.stat = c( "rsq", "f" ),
           omit = "cbsa" )

```

### Regression with multiple variables

```{r, results='asis', echo=FALSE}


m1 <- lm( mhv.change ~ p.white + p.col + cbsa, data=d.reg )
m2 <- lm( mhv.change ~ p.white + pov.rate + cbsa, data = d.reg)
m3 <- lm( mhv.change ~ pov.rate + p.col + cbsa, data = d.reg )
m4 <- lm( mhv.change ~ p.white + p.col + pov.rate + cbsa, data=d.reg)

stargazer( m1, m2, m3, m4,  
           type= "html", digits=2,
           omit.stat = c("rsq","f"),
           omit="cbsa" )

```


## Part 4 Results

Each of the three variables I chose (percent white residents, percent with college
or greater education level, and poverty rate) did significantly predict change in
median home values from 2000 to 2010. Controlling for metro area fixed effects, the 
amount of variance accounted for was approximately equal whether looking at any 
of these variables alone or in a model including two or three variable--each of
these models had an Adjusted R2 of 0.29 to 0.31. This is likely because of the strong
correlations between the variables. 

Model 4 with all three variables included produced an Adjusted R2 of 0.31, 
significant  at the p<0.01 level. Each of these variables had a negative correlation,
thus a higher percentage of white residents, higher percentage with a college
education, or higher poverty rate all indicate a smaller change in median home values.

I expected to see these relationships for percent of white and college-educated 
residents, since tracts with these demographics may already be high-priced and not
eligible for gentrification (and I also expect the converse, that areas with low 
white and college-educated populations may be ripe for gentrification). However, I
am surprised to see that poverty rate is also a negative correlation. This may be due 
to multicollinearity, since college education and poverty rate are closely correlated. 
In Model 2 where college education is not included, poverty rate has a positive
correlation with median home value change.







