---
title: "CPP 528 Lab 05"
author: "Caitlyn McCullers"
date: "4/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library( tidyverse )
library( stargazer )
library( dplyr )
library( here )
library( knitr )
library( pander )
library( scales )
# set randomization seed ----
set.seed( 1234 )
```


```{r, results = 'asis', echo=FALSE}
# load necessary functions and objects ----
# note: all of these are R objects that will be used throughout this .rmd file
import::here("S_TYPE",
             "d",
             "df",
             "d3",
             "d6",
             "PLOTS",
             "%>%",
             # notice the use of here::here() that points to the .R file
             # where all these R objects are created
             .from = here::here("labs/wk05/lab_05_source.R"),
             .character_only = TRUE)
```



## Difference in Difference Regression for NMTC without controls

<br>

```{r, results = 'asis', echo=FALSE}
m.nmtc <- lm( y ~ treat.nmtc + post + treat.nmtc * post, data = d3 )
stargazer::stargazer(m.nmtc,
                     type = "html",
                     digits = 2 )
```
<br>
<br>

#### Calculating results (NMTC no controls)

<br>

C1 = first timepoint, no treatment
```{r, echo=FALSE}
b0 <- m.nmtc$coefficients[1] 
b1 <- m.nmtc$coefficients[2]
b2 <- m.nmtc$coefficients[3]
b3 <- m.nmtc$coefficients[4]
# C1 = B0  
exp( b0 )
```
C2 = second timepoint, no treatment
```{r, echo=FALSE}
# C2 = B0 + B2 
exp( b0 + b2 )
```
T1 = first timepoint, intervention group
```{r, echo=FALSE}
# T1 = B0 + B1 
exp( b0 + b1 )
```
T2 = second timepoint (post-intervention), intervention group
```{r, echo=FALSE}
# T2 = B0 + B1 + B2 + B3 
exp( b0+b1+b2+b3 )
```
<br>

### NMTC results without including control variables

C1 = 156581.4 <br>
C2 = 196984.4 <br>
T1 = 120265.3 <br>
T2 = 166560.5 <br>

In this model without any control variables added, New Market Tax Credit investments 
had a significant effect on median home values, with the value of homes in tracts 
that received NMTC investments increasing less than those in tracts without NMTC.
The median house value was $5892.20 less in 2010 data than would be expected if 
NMTC investments were not made. 
However, because NMTC are only allowed in tracts identified because of poverty or
low family incomes, this finding seems likely to be the result of confounding
variables such as poverty.


## NMTC regression with control variables added

Variables of interest:

* Percent white residents in 2000 (p.white.00)
* Percent with 4-year college education or more (p.col.edu.00)
* Poverty rate in 2000 (pov.rate.00)

<br>

```{r, results = 'asis', echo=FALSE}
m.nmtc.controls <- lm( y ~ treat.nmtc + post +treat.nmtc * post + p.white + 
                         p.col.edu.00 + pov.rate.00, data = d3 )
stargazer::stargazer(m.nmtc.controls,
                     type = "html",
                     digits = 2 )
```
<br>
<br>

#### Calculating results (NMTC *with* controls)

<br>

C1
```{r, echo=FALSE}
b0.nmtc <- m.nmtc.controls$coefficients[1] 
b1.nmtc <- m.nmtc.controls$coefficients[2]
b2.nmtc <- m.nmtc.controls$coefficients[3]
b3.nmtc <- m.nmtc.controls$coefficients[4]
# C1 = B0  
exp( b0.nmtc )
```
C2
```{r, echo=FALSE}
# C2 = B0 + B2 
exp( b0.nmtc + b2.nmtc )
```
T1
```{r, echo=FALSE}
# T1 = B0 + B1 
exp( b0 + b1 )
```
T2
```{r, echo=FALSE}
# T2 = B0 + B1 + B2 + B3 
exp( b0+b1+b2+b3 )
```
<br>

### NMTC results with control variables

Explanation:
When percent white residents, percent of college educated residents, and poverty 
rate in 2000 are included as controls, New Market Tax Credits had a significant
effect on median home value. New Market Tax Credits resulted in a median home 
value of $166,560.50, an increase of $7681 over the expected value if New Market 
Tax Credits had not been utilized.



## Difference in Difference Regression for LIHTC without controls

```{r, results = 'asis', echo=FALSE}
m.lihtc <- lm( y ~ treat.lihtc + post + treat.lihtc * post, data = d6 )
stargazer::stargazer(m.lihtc,
                     type = "html",
                     digits = 2 )
```

<br>
<br>

#### Calculating results (LIHTC no controls)

<br>

C1
```{r, echo=FALSE}
b0 <- m.lihtc$coefficients[1] 
b1 <- m.lihtc$coefficients[2]
b2 <- m.lihtc$coefficients[3]
b3 <- m.lihtc$coefficients[4]
# C1 = B0  
exp( b0 )
```
C2
```{r, echo=FALSE}
# C2 = B0 + B2 
exp( b0 + b2 )
```
T1
```{r, echo=FALSE}
# T1 = B0 + B1 
exp( b0 + b1 )
```
T2
```{r, echo=FALSE}
# T2 = B0 + B1 + B2 + B3 
exp( b0+b1+b2+b3 )
```
<br>

### LIHTC results without control variables

In this model, which does not include any control variables, the Low Income Housing
Tax Credit program did not have a significant effect on median home values.


## LIHTC Regression with control variables added

Variables of interest:

* Percent white residents in 2000 (p.white.00)
* Percent with 4-year college education or more (p.col.edu.00)
* Poverty rate in 2000 (pov.rate.00)


```{r, results = 'asis', echo=FALSE}
m.lihtc.controls <- lm( y ~ treat.lihtc + post +treat.lihtc * post + p.white + 
                         p.col.edu.00 + pov.rate.00, data = d6 )
stargazer::stargazer(m.lihtc.controls,
                     type = "html",
                     digits = 2 )
```
<br>
<br>

#### Calculating results (LIHTC *with* controls)

<br>
C1
```{r, echo=FALSE}
b0 <- m.lihtc.controls$coefficients[1] 
b1 <- m.lihtc.controls$coefficients[2]
b2 <- m.lihtc.controls$coefficients[3]
b3 <- m.lihtc.controls$coefficients[4]
# C1 = B0  
exp( b0 )
```
C2
```{r, echo=FALSE}
# C2 = B0 + B2 
exp( b0 + b2 )
```
T1
```{r, echo=FALSE}
# T1 = B0 + B1 
exp( b0 + b1 )
```
T2
```{r, echo=FALSE}
# T2 = B0 + B1 + B2 + B3 
exp( b0+b1+b2+b3 )
```
<br>

### LIHTC results with control variables

When percent white residents, percent of college educated residents, and poverty 
rate in 2000 are included as controls, Low Income Housing Tax Credits did not have
a significant effect on median home values. In this model, median home price is
nearly identical in tracts with or without LIHTC. 


### Reflection: How can we test the parallel lines assumption in this model?

Examining longitudinal data from points before and/or after the timeframe in 
question would provide evidence that the parallel trends assumption is met, 
that both the treatment and control groups would have changed at the same rate 
in the absence of the treatment. 


<br>
<br>
