---
title: "Lab-03"
author: "Mohamed Said"
date: '2022-04-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
library( dplyr )
library( here )
library( knitr )
library( pander )
library( stargazer )
library( scales )

# set stargazer type to text for 
# previewing in RMD docs but
# convert to type HTML when knitting
# (next code chunk)

library( tidycensus )
library( ggthemes )
library( shinyjs )
library(import)
library( geojsonio )  # read geoJSON map files from GitHub
library( sp )         # spatial data class sp for shapefiles
library( cartogram )  # spatial maps w/ tract size bias reduction
library( tmap )       # thematic maps
library( maptools )   # spatial object manipulation 
library( sf )         # 'simple features' flavor of shapefiles
library( ggplot2 )     # graphing 
library( ggthemes )    # nice formats for ggplots
library( dplyr )       # data wrangling 
library( pander )      # formatting RMD tables


key <- "f2839493bf1c94e411da4713432bc0ee88e0ed4b"
census_api_key( key )
# change to s.type <- "html" before knitting
s.type <- "html"  
```

#Part 01 - Change in Home Values

```{r}
# import specific functions
# note: all of these are R objects that will be used throughout this .rmd file
import::here("clean_d",
             "tidy_up_data",
             "build_year",
             "RELEVANT_FILES",
             "obtain_crosswalk",
             "create_final_metadata_file",
             # notice the use of here::here() that points to the .R file
             # where all these R objects are created
             .from = here::here("labs/wk03/utilities.R"),
             .character_only = TRUE)
```


```{r}
# note: please do not use static file paths
# note: notice down below the use of here::here()
d1 <- readRDS( here::here( "data/rodeo/LTDB-2000.rds" ) )
d2 <- readRDS( here::here( "data/rodeo/LTDB-2010.rds" ) )
md <- readRDS( here::here( "data/rodeo/LTDB-META-DATA.rds" ) )

# check to make sure we are not losing 
# or gaining observations in the merge
nrow( d1 )
```

```{r}
```{r}
d1 <- select( d1, - year )
d2 <- select( d2, - year )
d <- merge( d1, d2, by="tractid" )
d <- merge( d, md, by="tractid" )
nrow( d )
```


# Filter Rural Districts
```{r}
table( d$urban )

d <- filter( d, urban == "urban" 
```
#Identify Common Variables

```{r}
find variables that are in both files
compare_dfs <- function( df1, df2 )
{
  # use regular expressions to remove numeric suffixes 
  var.names.1 <- names( df1 )
  var.names.1 <- gsub( "[.][xy]$", "", var.names.1 )
  var.names.1 <- gsub( "[0-9]{2}$", "", var.names.1 )
  
  var.names.2 <- names( df2 )
  var.names.2 <- gsub( "[.][xy]$", "", var.names.2 )
  var.names.2 <- gsub( "[0-9]{2}$", "", var.names.2 )
  
  shared <- intersect( var.names.1, var.names.2 ) %>% sort()
  print( "SHARED VARIABLES:")
  print( shared )
  
  not.shared <- c( setdiff( var.names.1, var.names.2 ),
                   setdiff( var.names.2, var.names.1 ) ) %>% sort()
  
  print( "NOT SHARED:" )
  print( not.shared )
  
  d.vars1 <- data.frame( type="shared", variables=shared, stringsAsFactors=F )
  d.vars2 <- data.frame( type="not shared", variables=not.shared, stringsAsFactors=F )
  dd <- rbind( d.vars1, d.vars2 )
  
  return( dd )
}

vars <- compare_dfs( df1=d1, df2=d2 )
```


```{r}
head( vars )

```

# Create Dataset for Analysis


```{r}

```{r}
d.full <- d  # keep a copy so don't have to reload 
```

# 1990-2000 variables

```{r}
d <- d.full  # story original in case you need to reset anything
d <- select( d, tractid, mhmval90, mhmval00, hinc90, hinc00, 
             hu90, own90, rent90,  
             empclf90, clf90, unemp90, prof90,  
             dpov90, npov90,
             ag25up90, hs90, col90, 
             pop90.x, nhwht90, nhblk90, hisp90, asian90,  ## ? pop90 or pop90.x
             cbsa, cbsaname )
d <- 
  d %>%
  mutate( p.white = 100 * nhwht90 / pop90.x,
          p.black = 100 * nhblk90 / pop90.x,
          p.hisp = 100 * hisp90 / pop90.x, 
          p.asian = 100 * asian90 / pop90.x,
          p.hs = 100 * (hs90+col90) / ag25up90,
          p.col = 100 * col90 / ag25up90,
          p.prof = 100 * prof90 / empclf90,
          p.unemp = 100 * unemp90 / clf90,
          pov.rate = 100 * npov90 / dpov90 )
```


```{r}
stargazer( d,
           type = s.type,
           digits = 0, 
           summary.stat = c( "min", "p25", "median", "mean", "p75", "max" ) )
```


#Exploration of median home value


#calculate inflation 1990 to 2000, $1 in 1990 was worth $1.32 in 2000 (from https://www.minneapolisfed.org/about-us/monetary-policy/inflation-calculator)


```{r, results='asis'}
# 1990-2000 inflation
# adjust 1990 home values for inflation 
mhv.90 <- d$mhmval90 * 1.31
mhv.00 <- d$mhmval00
mhv.change <- mhv.00 - mhv.90
df <- data.frame( MedianHomeValue1990=mhv.90, 
                  MedianHomeValue2000=mhv.00, 
                  Change.90.to.00=mhv.change )
stargazer( df, 
           type=s.type, 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```
# Histogram of MHV

```{r}
hist( mhv.change/1000, breaks=500, 
      xlim=c(-100,500), yaxt="n", xaxt="n",
      xlab="Thousand of US Dollars (adjusted to 2000)", cex.lab=1.5,
      ylab="", main="Change in Median Home Value 1990 to 2000",
      col="gray20", border="white" )
axis( side=1, at=seq( from=-100, to=500, by=100 ), 
      labels=paste0( "$", seq( from=-100, to=500, by=100 ), "k" ) )
        
mean.x <- mean( mhv.change/1000, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=200, y=1500, 
      labels=paste0( "Mean = ", dollar( round(1000*mean.x,0)) ), 
      col="darkorange", cex=1.8, pos=3 )
median.x <- median( mhv.change/1000, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=200, y=2000, 
      labels=paste0( "Median = ", dollar( round(1000*median.x,0)) ), 
      col="dodgerblue", cex=1.8, pos=3 )
```


```{r}
 
# function to control plot() formatting 
jplot <- function( x1, x2, lab1="", lab2="", draw.line=T, ... )
{
    plot( x1, x2,
          pch=19, 
          col=gray(0.6, alpha = 0.2), 
          cex=2.5,  
          bty = "n",
          xlab=lab1, 
          ylab=lab2, cex.lab=1.5,
        ... )
    if( draw.line==T ){ 
        ok <- is.finite(x1) & is.finite(x2)
        lines( lowess(x2[ok]~x1[ok]), col="red", lwd=3 ) }
}
```



```{r}
layout.matrix <- matrix( c( 1,3,
                            2,3 ), 
                nrow=2, ncol=2, byrow=T )
layout( mat = layout.matrix,
        heights = c(2,2), # Heights of the two rows
        widths =  c(3,4)) # Widths of the two columns
# layout.show(3)
par( mar=c(4,0,0,2) )
hist( mhv.90/1000, breaks=50, 
      xlim=c(-200,800), yaxt="n", xaxt="n",
      xlab="", cex.lab=1,
      ylab="", main="",
      col="darkslateblue", border="white" )
axis( side=1, at=seq( from=0, to=1000, by=100 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=100 ), "k" ) )
abline( v=seq(0,1000,100), lty=2, col="gray80" )
text( 550, 4000, labels="Median Home \nValue in 1990", 
      col="darkslateblue", cex=1.8 )
hist( mhv.00/1000, breaks=50, 
      xlim=c(-200,800), yaxt="n", xaxt="n",
      xlab="", cex.lab=1,
      ylab="", main="",
      col="darkslateblue", border="white" )
abline( v=seq(0,1000, 100 ), lty=2, col="gray80" )
text( 550, 3500, labels="Median Home \nValue in 2000", 
      col="darkslateblue", cex=1.8 )
axis( side=1, at=seq( from=0, to=1000, by=100 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=100 ), "k" ) )
# data reduction - filter 1,000 observations
df <- data.frame( v00=mhv.90/1000, v10=mhv.00/1000 )
df <- sample_n( df, 1000 )
par( mar=c(4,5,3,2) )
jplot( df$v00, df$v10, 
       lab1="MHV in 1990", lab2="MHV in 2000",
       xlim=c(0,1000), ylim=c(0,1000),
       axes=F )
abline( a=0, b=1, lty=2, col="gray" )
axis( side=1, at=seq( from=0, to=1000, by=200 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=200 ), "k" ) )
axis( side=2, at=seq( from=0, to=1000, by=200 ), 
      labels=paste0( "$", seq( from=0, to=1000, by=200 ), "k" ) )
```


# Change in MHV 1990-2000

```{r}
# quick filter to remove all of the problematic obs
# but need to go back and see which cases are problematic
mhv.90[ mhv.90 < 10000 ] <- NA
pct.change <- mhv.change / mhv.90
summary( pct.change )
```



```{r}
# how many cases had increases above 500%
sum( pct.change > 5, na.rm=T )
```



```{r}
# preview tracts with large increases in home values 
# to see if increases make sense 
d %>% 
  filter( pct.change > 5 ) %>% 
  head()
```


# Plot the percent change variable
```{r}
hg <-
hist( pct.change, breaks=2000, 
      xlim=c(-1,2), yaxt="n", xaxt="n",
      xlab="", cex.main=1.5,
      ylab="", main="Growth in Home Value by Census Tract 1990 to 2000",
      col="gray40", border="white" )
axis( side=1, at=seq( from=-1, to=2, by=0.5 ), 
      labels=paste0( seq( from=-100, to=200, by=50 ), "%" ) )
ymax <- max( hg$count )
        
mean.x <- mean( pct.change, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=1, y=(0.5*ymax), 
      labels=paste0( "Mean = ", round(100*mean.x,0), "%"), 
      col="darkorange", cex=1.8, pos=4 )
median.x <- median( pct.change, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=1, y=(0.6*ymax), 
      labels=paste0( "Median = ", round(100*median.x,0), "%"), 
      col="dodgerblue", cex=1.8, pos=4 )
```

# group growth rates by metro area

```{r}
d$mhv.change <- mhv.change 
d$pct.change <- pct.change
d$mhv.00 <- mhv.00
d$mhv.90 <- mhv.90
d %>%
  group_by( cbsaname ) %>%
  summarize( ave.change = median( mhv.change, na.rm=T ),
             ave.change.d = dollar( round(ave.change,0) ),
             growth = 100 * median( pct.change, na.rm=T ) ) %>%
  ungroup() %>%
  arrange( - growth ) %>%
  select( - ave.change ) %>% 
  head( 25 ) %>%
  pander()
```


# Part one solution:
# From 1990-2000 the median home value increased from $113,446 to $119,900, while the mean value decreased from $147,601 to $144,738 during the same period. The 2000-2010 period saw an increase from $119,000 to 193,200 as well as a dramatic mean increase from $144,738 to $246,570.




# Part 02 - Measuring Gentrification

```{r}
# adjust 1990 home values for inflation 
mhv.90 <- d.full$mhmval90 * 1.31  
mhv.00 <- d.full$mhmval00
mhv.change <- mhv.00 - mhv.90
# small initial values are skewing percentages
#
# an average home value below $10k is really low -
# these must be mostly vacant lots?
mhv.90[ mhv.90 < 10000 ] <- NA
pct.change <- 100 * ( mhv.change / mhv.90 )
summary( pct.change )
```



```{r}
d.full$mhv.90 <- mhv.90
d.full$mhv.00 <- mhv.00
d.full$mhv.change <- mhv.change
d.full$pct.change <- pct.change
```


# Select Gentrification Variables

```{r}
head( vars )
```

```{r}
``{r}
d3 <- select( d.full, 
             
             tractid, cbsa, cbsaname,            # ids / units of analysis
             
             mhv.90, mhv.00, mhv.change, pct.change,    # home value 
             
             mrent90, mrent00,                   # rent
             
             vac90, vac00,                       # vacant units
             
             hinc90, hu90, own90, rent90,        # ses
             hinc00, hu00, own00, rent00,
             
             empclf90, clf90, unemp90, prof90,   # employment 
             empclf00, clf00, unemp00, prof00,
             
             dpov90, npov90,                     # poverty
             dpov00, npov00,
             
             ag25up90, hs90, col90,              # education 
             ag25up00, hs00, col00,
             
             pop90.x, nhwht90, nhblk90, hisp90, asian90,   # race
             pop00.x, nhwht00, nhblk00, hisp00, asian00
             
          ) # end select
d3 <- 
  d3 %>%
  mutate( 
          # 1990 variables
          p.white.90 = 100 * nhwht90 / pop90.x,
          p.black.90 = 100 * nhblk90 / pop90.x,
          p.hisp.90 = 100 * hisp90 / pop90.x, 
          p.asian.90 = 100 * asian90 / pop90.x,
          p.hs.edu.90 = 100 * (hs90+col90) / ag25up90,
          p.col.edu.90 = 100 * col90 / ag25up90,
          p.prof.90 = 100 * prof90 / empclf90,
          p.unemp.90 = 100 * unemp90 / clf90,
          pov.rate.90 = 100 * npov90 / dpov90,
          p.vac.90 = 100 * vac90 / hu90,
          
          # 2000 variables
          p.white.00 = 100 * nhwht00 / pop00.x,
          p.black.00 = 100 * nhblk00 / pop00.x,
          p.hisp.00 = 100 * hisp00 / pop00.x, 
          p.asian.00 = 100 * asian00 / pop00.x,
          p.hs.edu.00 = 100 * (hs00+col00) / ag25up00,
          p.col.edu.00 = 100 * col00 / ag25up00,
          p.prof.00 = 100 * prof00 / empclf00,
          p.unemp.00 = 100 * unemp00 / clf00,
          pov.rate.00 = 100 * npov00 / dpov00, 
          p.vac.00 = 100 * vac00 / hu00
)
          
```


```{r}
d3 <-
  d3 %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.pct.90 = ntile( mhv.90, 100 ),
          metro.mhv.pct.00 = ntile( mhv.00, 100 ),
          metro.median.pay.90 = median( hinc90, na.rm=T ),
          metro.median.pay.00 = median( hinc00, na.rm=T ),
          metro.race.rank.90 = ntile( (100-p.white.90), 100 ) ) %>%
  ungroup() %>%
  mutate( metro.mhv.pct.change = metro.mhv.pct.00 - metro.mhv.pct.90,
          pay.change = metro.median.pay.00 - metro.median.pay.90,
          race.change = p.white.00 - p.white.90,
          mhv.change = mhv.00 - mhv.90 )
```

# Descriptive Statistics of Change Variables

```{r}
d3 <-           
  d3 %>%
  select( c( "tractid", "cbsa", "cbsaname",
             "mhv.90", "mhv.00", "mhv.change","pct.change",
          "p.white.90", "p.black.90", "p.hisp.90", "p.asian.90", 
          "p.hs.edu.90", "p.col.edu.90", "p.prof.90",  "p.unemp.90", 
          "pov.rate.90", "mrent90", "p.vac.90", "p.white.00", "p.black.00", "p.hisp.00", 
          "p.asian.00", "p.hs.edu.00", "p.col.edu.00", "p.prof.00", 
          "p.unemp.00", "pov.rate.00", "metro.mhv.pct.90", 
          "metro.mhv.pct.00", "metro.median.pay.90", "metro.median.pay.00", 
          "metro.mhv.pct.change", "pay.change", "race.change",
          "metro.race.rank.90") ) 
```

```{r, results='asis'}
d3 <- data.frame(d3)
stargazer( d3, 
           type=s.type, 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

# Operationalizing Gentrification; Based on proposal by Sarah J. to be used by group, the following factor will pridct the tract gentrificaion  (- Low home values, - Low rents,- High rate of vacant units,  - High proportion of racial/ethnic minority residents, - High rate of unemployment, - Low % with college degree or more education and - High poverty rate)
```{r}
## Predicted factors that make a tract eligible for gentrification
# home value in lower than average home in a metro in 1990
lowhomeval.1990 <- d3$metro.mhv.pct.90 < 50  
# vacancies above 75th percentile
vacant.1990 <- d3$p.vac.90 > 9
# median rent below the 25th percentile
lowrent.1990 <- d3$mrent90 < 168
# above average diversity for metro
diverse.1990 <- d3$metro.race.rank.90 > 50 
# unemployment above 75th percentile
unemployment.1990 <- d3$p.unemp.90 > 8
# college or more education below the 25th percentile
lowed.1990 <- d3$p.col.edu.90 < 10
# poverty rate above 75h percentile
highpoverty.1990 <- d3$pov.rate.90 > 16
sum( lowhomeval.1990, na.rm = T )
sum( vacant.1990, na.rm = T )
sum( lowrent.1990, na.rm = T )
sum( diverse.1990, na.rm = T )
sum( unemployment.1990, na.rm = T )
sum( lowed.1990, na.rm = T )
sum( highpoverty.1990, na.rm = T )
```

```{r}
gent.candidate <- sum( lowhomeval.1990 & vacant.1990  & lowrent.1990 & 
                         diverse.1990 & unemployment.1990 & lowed.1990 & 
                         highpoverty.1990, na.rm = T )
gent.candidate
```
# Using all of the criteria listed, only 418 tracts were eligible for gentrification

```{r}
gent.candidate2 <- sum(( lowhomeval.1990 | vacant.1990 | lowrent.1990 ) & 
                          diverse.1990 & (unemployment.1990 | highpoverty.1990 ) &
                          lowed.1990, na.rm = T )
gent.candidate2
```

# By combining the three property criteria (so a tract can be selected if it has low home values and/or high vacancies and/or low rent), and combining the two income factors ( unemployment and/or high poverty rate ), the number of tracts eligible for gentrification increased 7315.



```{r}
``{r}
# home values increased more than overall city gains 
# change in percentile rank within the metro
mhv.pct.increase <- d3$metro.mhv.pct.change > 0
# faster than average growth  
# 25% growth in value is median for the country
home.val.rise <- d3$pct.change > 25 
# proportion of whites increases by more than 3 percent 
# measured by increase in white
loss.diversity <- d3$race.change > 3 
```


```{r}
gentrified <- ( lowhomeval.1990 | vacant.1990 | lowrent.1990 ) & 
                          diverse.1990 & (unemployment.1990 | highpoverty.1990 ) &
                          lowed.1990 & mhv.pct.increase & home.val.rise & loss.diversity
num.gentrified <- sum( gentrified, na.rm=T )
num.gentrified 
gent.candidate2
pct.gentrified <- num.gentrified / gent.candidate2
pct.gentrified
```

#Using this approach, only 1.85% of eligible tracts became gentrified from 1990 to 2000.

```{r}
# small initial values are skewing percentages
#
# an average home value below $10k is really low -
# these must be mostly vacant lots?
mhv.90[ mhv.90 < 1000 ] <- NA
pct.change <- 100 * ( mhv.change / mhv.90 )
summary( pct.change )
```


# Part 03 - Spatial Patterns

```{r}
crosswalk <- read.csv( "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/cbsatocountycrosswalk.csv",  stringsAsFactors=F, colClasses="character" )
# search for city names by strings, use the ^ anchor for "begins with" 
grep( "^MIL", crosswalk$msaname, value=TRUE ) 
```


```{r}
these.MKE <- crosswalk$msaname == "MILWAUKEE-WAUKESHA, WI"
these.fips <- crosswalk$fipscounty[ these.MKE ]
these.fips <- na.omit( these.fips )
state.fips <- substr( these.fips, 1, 2 )
county.fips <- substr( these.fips, 3, 5 )
MKE <-
get_acs( geography = "tract", variables = "B01003_001",
         state = "55", county = county.fips[state.fips=="55"], 
         geometry = TRUE, progress_bar = FALSE ) %>% 
         select( GEOID, estimate ) %>%
         rename( POP=estimate )
```


```{r}
plot( MKE)
```


```{r}
# create small dataframe for the merge
df <- data.frame(  tractid=d$tractid, 
        mhv.90, mhv.00,  mhv.change,  pct.change, loss.diversity, lowhomeval.1990,
        unemployment.1990, d3$p.unemp.90, d3$p.unemp.00, d3$pov.rate.90, d3$pov.rate.00, d3$pay.change, 
        d3$race.change )
# create GEOID that matches GIS format
# create a geoID for merging by tract 
df$GEOID <- substr( df$tractid, 6, 18 )  # extract codes
df$GEOID <- gsub( "-", "", df$GEOID )    # remove hyphens
class( df$GEOID )
```


```{r}
head( df$GEOID )
```

```{r}
nrow( MKE )
```


```{r}
mke <- merge( MKE, df, by.x="GEOID", by.y="GEOID" )
nrow( mke )
```


```{r}
mke <- as_Spatial( mke )
mke <- spTransform( mke, CRS("+init=epsg:3395"))
mke_dorling <- cartogram_dorling( x=mke, weight = "mhv.00", k=0.05 )
plot( mke_dorling )
tm_shape( mke_dorling ) + 
  tm_polygons( col="mhv.90", n=10, style="quantile", palette="Spectral" ) +
  tm_layout( "Dorling Cartogram", title.position=c("right","top") )
```


### Home Values

**Describe the distribution of home values in 1990 - where are high and low-value tracts located in the city/cities?**

#In 1990, high-value tracts were concentrated north of downtown along the lakefront (the far right edge of the map) and scattered through the northern and western suburbs. Low-value tracts are located in the central city, southern city and suburbs, and northwest areas of the city.

```{r}
tm_shape( mke_dorling) + 
  tm_polygons( col="mhv.change", n=10, style="quantile", palette="Spectral" ) +
  tm_layout( "Dorling Cartogram", title.position=c("right","top") )
```
**Compare values in 2000 to changes in values from 1990-2000. Do the largest gains**
**occur in tracts with above or below-average home prices in 2000?**
# The most striking areas with gains from 1990-2000 were lower-value but adjacent to high-value areas in 1990. There is a cluster with gains downtown and just west of the lakefront, while areas along the lake downtown had losses during this period.

```{r}
tm_shape( mke_dorling ) + 
  tm_polygons( col="pct.change", n=10, style="quantile", palette="Spectral" ) +
  tm_layout( "Dorling Cartogram", title.position=c("right","top") )
```


```{r}
tm_shape( mke_dorling ) + 
  tm_polygons( col="d3.pov.rate.90", n=10, style="quantile", palette="Spectral" ) +
  tm_layout( "Dorling Cartogram", title.position=c("right","top") )
tm_shape( mke_dorling ) + 
  tm_polygons( col="d3.pov.rate.00", n=10, style="quantile", palette="Spectral" ) +
  tm_layout( "Dorling Cartogram", title.position=c("right","top") )
tm_shape( mke_dorling ) + 
  tm_polygons( col="d3.race.change", n=10, style="quantile", palette="Spectral" ) +
  tm_layout( "Dorling Cartogram", title.position=c("right","top") )
```

#These maps along with the map above showing percent change in median home value change from 1990 to 2000 help show the areas where in 1990 there was very high poverty and between 1990 and 2000 property values rose at the same time that the racial mix of residents stayed fairly stable or became more white. It does appear that the area just west of the lakefront in the downtown area experienced gentrification during this time period. 


